
# ATLANTICMACKEREL

## Map

```{r}
url <- here::here("data-raw/2026/ATLANTICMACKEREL_map.png")
```

```{r, eval = file.exists(url)}
knitr::include_graphics(url)
```

```{r, eval = !file.exists(url)}
cat("The map for this species hasn't been created yet.")
```

## Socioeconomic indicators

```{r socioeco-data}
url <- here::here("data/intermediate/ATLANTICMACKEREL_Commercial_Indicators_Master.csv")
data <- read.csv(url)
```

Get the data used to make the socioeconomic indicator plots [here](`r url`).

### Average price per pound

```{r}
avgprice <- data |>
  dplyr::filter(INDICATOR_NAME == "AVGPRICE_ATLANTICMACKEREL_2025_DOLlb") 

NEesp2::plt_indicator(data = avgprice,
                      ar = 1/4,
                      include_trends = TRUE) |>
  plotly::ggplotly() 
```

### Total commercial landings (pounds)

```{r}
landings <- data |>
  dplyr::filter(INDICATOR_NAME == "Commercial_ATLANTICMACKEREL_Landings_LBS") 

NEesp2::plt_indicator(data = landings,
                      ar = 1/4,
                      include_trends = TRUE) |>
  plotly::ggplotly() 
```

### Total commercial revenue

```{r}
totalrev <- data |>
  dplyr::filter(INDICATOR_NAME == "TOTALANNUALREV_ATLANTICMACKEREL_2025Dols") 

NEesp2::plt_indicator(data = totalrev,
                      ar = 1/4,
                      include_trends = TRUE) |>
  plotly::ggplotly() 
```

### Number of commercial vessels

```{r}
nvessels <- data |>
  dplyr::filter(INDICATOR_NAME == "N_Commercial_Vessels_Landing_ATLANTICMACKEREL") 

NEesp2::plt_indicator(data = nvessels,
                      ar = 1/4,
                      include_trends = TRUE) |>
  plotly::ggplotly() 
```

### Average revenue per vessel ($USD2024)

```{r}
vesselrev <- data |>
  dplyr::filter(INDICATOR_NAME == "AVGVESREVperYr_ATLANTICMACKEREL_2025_DOLlb") 

NEesp2::plt_indicator(data = vesselrev,
                      ar = 1/4,
                      include_trends = TRUE) |>
  plotly::ggplotly() 
```

## Ecosystem indicators

### Monthly bottom temperature in stock area

```{r bt-data}
url1 <- here::here("data-raw/2026/ATLANTICMACKEREL_hubert_bottomT.csv")
url2 <- here::here("data-raw/2026/ATLANTICMACKEREL_glorys_bottomT.csv")
```


```{r, eval = file.exists(url1) & file.exists(url2), results = "asis"}
cat("Get the data used to make the bottom temperature indicator plots [here (pre-2021)](", url1, ") and [here (2021 and later)](", url2, ").", sep = "")
```

```{r, eval = file.exists(url1) & file.exists(url2), results='asis'}
bt_data <- read.csv(url1) |>
  dplyr::bind_rows(read.csv(url2)) |>
  dplyr::group_by(AREA, MONTH, YEAR, INDICATOR_NAME, INDICATOR_UNITS) |>
  dplyr::summarise(DATA_VALUE = mean(DATA_VALUE, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(INDICATOR_NAME = "monthly_bottomT")

for(i in 1:12) {
  print(
    bt_data |>
      dplyr::filter(MONTH == i) |>
      NEesp2::plt_indicator(
        ar = 1/4,
        include_trends = TRUE
      ) +
      ggplot2::ggtitle(paste0("Monthly bottom temperature in stock area - ", month.abb[i]))
  )
}
# 
# for(i in 1:12) {
#   knitr::knit_child(text = paste("```{r}\nknitr::knit_child(text =  bt_data |>
#       plt <- dplyr::filter(MONTH ==", i,") |>
#       NEesp2::plt_indicator(
#         ar = 1/4,
#         include_trends = TRUE
#       ) |>
#   plotly::ggplotly() |>
#   plotly::layout(title = list(text = paste0('Monthly bottom temperature in stock area - ', month.abb[", i, "])))  
#   \nprint(plt)\n```"), quiet = TRUE)
# }

# for(i in 1:12) {
#   # Create the internal R code as a string first for readability
#   child_code <- paste0(
#     "```{r, echo=FALSE, message=FALSE}\n",
#     "plt <- bt_data |> \n",
#     "  dplyr::filter(MONTH == ", i, ") |> \n",
#     "  NEesp2::plt_indicator(ar = 1/4, include_trends = TRUE) |> \n",
#     "  plotly::ggplotly() |> \n",
#     "  plotly::layout(title = list(text = paste0('Monthly bottom temperature in stock area - ', month.abb[", i, "]))) \n",
#     "print(htmltools::tagList(plt)) \n",
#     "```"
#   )
#   
#   # Knit the child chunk
#   cat(knitr::knit_child(text = child_code, quiet = TRUE))
# }

# for(i in 1:12) {
#   print(
#     bt_data |>
#       dplyr::filter(MONTH == i) |>
#       NEesp2::plt_indicator(
#         ar = 1/4,
#         include_trends = TRUE
#       ) |>
#   plotly::ggplotly() |>
#   plotly::layout(title = list(text = paste0("Monthly bottom temperature in stock area - ", month.abb[i])))
#   )
# }


```

```{r, eval = !(file.exists(url1) & file.exists(url2))}
cat("Bottom temperature for this species hasn't been calculated yet.")
```

### Monthly bottom salinity in stock area


```{r bs-data}
 url3 <- here::here("data-raw/2026/ATLANTICMACKEREL_glorys_bottomS.csv")
```

```{r, eval = file.exists(url3), results = "asis"}
cat("Get the data used to make the bottom salinity indicator plots [here](", url3, ").")
```

```{r, eval = file.exists(url3)}
bs_data <- read.csv(url3) |>
  dplyr::group_by(AREA, MONTH, YEAR, INDICATOR_NAME, INDICATOR_UNITS) |>
  dplyr::summarise(DATA_VALUE = mean(DATA_VALUE, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(INDICATOR_NAME = "monthly_bottomS")

for(i in 1:12) {
  print(
    bs_data |>
      dplyr::filter(MONTH == i) |>
      NEesp2::plt_indicator(
        ar = 1/4,
        include_trends = TRUE
      ) +
      ggplot2::ggtitle(paste0("Monthly bottom salinity in stock area - ", month.abb[i]))
  )
}

# for(i in 1:12) {
#   print(
#     bs_data |>
#       dplyr::filter(MONTH == i) |>
#       NEesp2::plt_indicator(
#         ar = 1/4,
#         include_trends = TRUE
#       ) |>
#   plotly::ggplotly() |>
#   plotly::layout(title = list(text = paste0("Monthly bottom temperature in stock area - ", month.abb[i])))
#   )
# }
```

```{r, eval = !file.exists(url3)}
cat("Bottom salinity for this species hasn't been calculated yet.")
```

### Monthly mean chlorophyll in stock area


```{r chl}
 url4 <- here::here("data-raw/2026/ATLANTICMACKEREL_chl.csv")
```

```{r, eval = file.exists(url4), results = "asis"}
cat("Get the data used to make the chlorophyll indicator plots [here](", url4, ").")
```

```{r, eval = file.exists(url4)}
chl <- read.csv(url4) |>
  dplyr::group_by(AREA, MONTH, YEAR, INDICATOR_NAME, INDICATOR_UNITS) |>
  dplyr::summarise(DATA_VALUE = mean(DATA_VALUE, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(INDICATOR_NAME = "monthly_chl")

for(i in 1:12) {
  print(
    chl |>
      dplyr::filter(MONTH == i) |>
      NEesp2::plt_indicator(
        ar = 1/4,
        include_trends = TRUE
      ) +
      ggplot2::ggtitle(paste0("Monthly mean chlorophyll in stock area - ", month.abb[i]))
  )
}

# for(i in 1:12) {
#   print(
#     chl |>
#       dplyr::filter(MONTH == i) |>
#       NEesp2::plt_indicator(
#         ar = 1/4,
#         include_trends = TRUE
#       ) |>
#   plotly::ggplotly() |>
#   plotly::layout(title = list(text = paste0("Monthly bottom temperature in stock area - ", month.abb[i])))
#   )
# }
```

```{r, eval = !file.exists(url4)}
cat("Monthly mean chlorophyll for this species hasn't been calculated yet.")
```

### Monthly sea surface temperature in stock area


```{r sst}
 url5 <- here::here("data-raw/2026/ATLANTICMACKEREL_sst.csv")
```

```{r, eval = file.exists(url5), results = "asis"}
cat("Get the data used to make the sea surface temperature indicator plots [here](", url5, ").")
```

```{r, eval = file.exists(url5)}
sst <- read.csv(url5) |>
  dplyr::group_by(AREA, MONTH, YEAR, INDICATOR_NAME, INDICATOR_UNITS) |>
  dplyr::summarise(DATA_VALUE = mean(DATA_VALUE, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(INDICATOR_NAME = "monthly_sst")

for(i in 1:12) {
  print(
    sst |>
      dplyr::filter(MONTH == i) |>
      NEesp2::plt_indicator(
        ar = 1/4,
        include_trends = TRUE
      ) +
      ggplot2::ggtitle(paste0("Monthly sea surface temperature in stock area - ", month.abb[i]))
  )
}

```

```{r, eval = !file.exists(url5)}
cat("Monthly sea surface temperature for this species hasn't been calculated yet.")
```
