
# ATLANTICMACKEREL

## Map

```{r}
url <- here::here("data-raw/2026/ATLANTICMACKEREL_map.png")
#url <- here::here('data-raw/2026', 'ATLANTICMACKEREL_map.png')
```

```{r, eval = file.exists(url)}
knitr::include_graphics(url)
```

```{r, eval = !file.exists(url)}
cat("The map for this species hasn't been created yet.")
```

## Socioeconomic indicators

```{r socioeco-data}
url <- here::here("data/intermediate/ATLANTICMACKEREL_Commercial_Indicators_Master.csv")
data <- read.csv(url)
```

Get the data used to make the socioeconomic indicator plots [here](`r url`).

### Average price per pound

```{r}
avgprice <- data |>
  dplyr::filter(INDICATOR_NAME == "AVGPRICE_ATLANTICMACKEREL_2025_DOLlb") 

NEesp2::plt_indicator(data = avgprice,
                      ar = 1/4,
                      include_trends = TRUE) 
```

### Total commercial landings (pounds)

```{r}
landings <- data |>
  dplyr::filter(INDICATOR_NAME == "Commercial_ATLANTICMACKEREL_Landings_LBS") 

NEesp2::plt_indicator(data = landings,
                      ar = 1/4,
                      include_trends = TRUE)
```

### Total commercial revenue

```{r}
totalrev <- data |>
  dplyr::filter(INDICATOR_NAME == "TOTALANNUALREV_ATLANTICMACKEREL_2025Dols") 

NEesp2::plt_indicator(data = totalrev,
                      ar = 1/4,
                      include_trends = TRUE)
```

### Number of commercial vessels

```{r}
nvessels <- data |>
  dplyr::filter(INDICATOR_NAME == "N_Commercial_Vessels_Landing_ATLANTICMACKEREL") 

NEesp2::plt_indicator(data = nvessels,
                      ar = 1/4,
                      include_trends = TRUE)
```

### Average revenue per vessel ($USD2024)

```{r}
vesselrev <- data |>
  dplyr::filter(INDICATOR_NAME == "AVGVESREVperYr_ATLANTICMACKEREL_2025_DOLlb") 

NEesp2::plt_indicator(data = vesselrev,
                      ar = 1/4,
                      include_trends = TRUE)
```

## Ecosystem indicators

### Monthly bottom temperature in stock area

```{r bt-data}
url1 <- here::here("data-raw/2026/ATLANTICMACKEREL_hubert_bottomT.csv")
#url1 <- here::here('data-raw/2026', 'ATLANTICMACKEREL_hubert_bottomT.csv')
url2 <- here::here("data-raw/2026/ATLANTICMACKEREL_glorys_bottomT.csv")
#url2 <- here::here('data-raw/2026', 'ATLANTICMACKEREL_glorys_bottomT.csv')
```


```{r, eval = file.exists(url1) & file.exists(url2), results = "asis"}
cat("Get the data used to make the bottom temperature indicator plots [here (pre-2021)](", url1, ") and [here (2021 and later)](", url2, ").", sep = "")
```

```{r, eval = file.exists(url1) & file.exists(url2)}
bt_data <- read.csv(url1) |>
  dplyr::bind_rows(read.csv(url2)) |>
  dplyr::group_by(AREA, MONTH, YEAR, INDICATOR_NAME, INDICATOR_UNITS) |>
  dplyr::summarise(DATA_VALUE = mean(DATA_VALUE, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(INDICATOR_NAME = "monthly_bottomT")

for(i in 1:12) {
  print(
    bt_data |>
      dplyr::filter(MONTH == i) |>
      NEesp2::plt_indicator(
        ar = 1/4,
        include_trends = TRUE
      ) +
      ggplot2::ggtitle(paste0("Monthly bottom temperature in stock area - ", month.abb[i]))
  )
}
```

```{r, eval = !(file.exists(url1) & file.exists(url2))}
cat("Bottom temperature for this species hasn't been calculated yet.")
```

### Monthly bottom salinity in stock area


```{r bs-data}
 url3 <- here::here("data-raw/2026/ATLANTICMACKEREL_glorys_bottomS.csv")
```

```{r, eval = file.exists(url3), results = "asis"}
cat("Get the data used to make the bottom salinity indicator plots [here](", url3, ").")
```

```{r, eval = file.exists(url3)}
bs_data <- read.csv(url3) |>
  dplyr::group_by(AREA, MONTH, YEAR, INDICATOR_NAME, INDICATOR_UNITS) |>
  dplyr::summarise(DATA_VALUE = mean(DATA_VALUE, na.rm = TRUE), .groups = "drop") |>
  dplyr::mutate(INDICATOR_NAME = "monthly_bottomS")

for(i in 1:12) {
  print(
    bs_data |>
      dplyr::filter(MONTH == i) |>
      NEesp2::plt_indicator(
        ar = 1/4,
        include_trends = TRUE
      ) +
      ggplot2::ggtitle(paste0("Monthly bottom salinity in stock area - ", month.abb[i]))
  )
}
```

```{r, eval = !file.exists(url3)}
cat("Bottom salinity for this species hasn't been calculated yet.")
```
